Mechanistic Bayesian Forecasts of COVID19

### 2020-08-27 paper_2020-08-26.tex
big update

The COVID-19 pandemic emerged in late December 2019. In the first six months of the global outbreak, the US reported more cases and deaths than any other country in the world. Effective modeling of the course of the pandemic can help assist with public health resource planning, intervention efforts, and vaccine clinical trials. However, data quality during a pandemic suffers from under-reporting, delayed reporting due to reporting infrastructure issues, and limited testing. Traditionally, compartmental models have been used to model emerging pandemics, however, the basic compartmental framework is difficult to operationalize in a complex data environment. We propose a novel Bayesian compartmental model (MechBayes) that builds upon the classic compartmental framework of susceptible-exposed-infected-recovered (SEIR) model to operationalize Covid-19 in forecasting in real time. This includes non-parametric modeling of varying transmission rates, non-parametric modelling of case detection versus death detection, and finally joint observation likelihood on new case counts and new deaths. The model has been used to submit forecasts to the US Centers for Disease Control, through the COVID19-ForecsatHub repository organized by the Reich Lab. We examine the performance relative to a baseline model in addition to performing an ablation test of our extensions to the classic SEIR models. We demonstrate a significant gain in both point and probabilistic forecast scoring measures using MechBayes.
%\begin{itemize}
%\item Covid has affected XX people
%\item Forecasts useful for public health resource planning, intervention planning (vaccine trials), and disease burden
%\item Basic Mechanistic Models unable to capture complexities of real world disease epidemics due to 
%\begin{itemize}
%\item Complexities of interventions 
%\item Issues with testing and reporting
%\end{itemize}
%\item We propose a novel forecasting algorithm to overcome 
%\begin{itemize}
%\item Under-reporting of cases
%\item Time-varying interventions
%\end{itemize}
%\item Bayesian end to end estimation using both cases and deaths in numpyro
%\end{itemize}

@@

In this work, we introduce a novel operational forecast model based on mechanistic foundations and tailored to the particular needs and data availability of COVID-19. These include, but are not limited to, severe under reporting of cases due to low testing rates especially in mild or asymptomatic cases, time-varying testing rates, delayed reporting data dumps, and both the addition and removal of control measures such as social distancing, lockdown, and mask use. The case and death data is reported as new cases, instead of prevalent, and new deaths, instead of cumulative, which we handle using an observation model on top of the underlying SEIR dynamics. Finally, we choose a Bayesian framework that allows for uncertainty in the epidemiological parameters that are unidentifiable from the data, introducing flexibility suited to forecasting. We distinguish this from scenario projection models, which require well identified epidemiological parameters that can be set to counterfactual values under different scenarios, such as an increase or decrease in intervention levels. Implementation in a Bayesian framework also allows for use of a cutting edge probabilistic programming framework for speed and rapid model development.

Our main goal in developing MechBayes is to forecast observed incident deaths, and because we have limited historical data on COVID-19, we think compartmental models are among the most parsimonious models available for forecasting. Our focus is not on identifying internal parameters of the model, many of which are poorly determined or not identifiable from the available data. We are willing to accept biologically unrealistic parameter values or priors for latent variables; but not ones that lead to gross pathologies during inference and forecasting. The non-parametric extensions to the underlying SEIR model also mean that any inference on epidemiological parameters should be carried out very carefully.

 We demonstrate the success of the model in both forecast submissions submitted to the US Centers for Disease Control via the COVID-19 Forecast Hub as well as an ablation model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model. In what follows we first describe the available data and forecast submission infrastructure, outline the basic SEIR compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.

@@

\section{Data}
	In this analysis we used confirmed case counts and deaths as reported by the Johns Hopkins University Center for Systems Science and Engineering \cite{dong2020interactive}. This a time series dataset which we truncate to begin March 1st 2020 to August 1st 2020 and captures all 50 states, as well as Guam, Puerto Rico, American Samoa, District of Columbia, Northern Mariana Islands and U.S. Virgin Islands. As noted in \cite{krantz2020level}, COVID-19 cases are under-reported, with the fraction of all infections reported as cases for the U.S. estimated at 20-30\% \cite{russel2020using}. 
		There are large discrepancies in reporting practices across the states (Figure \ref{fig:data}). For example, New Jersey reported an additional 1600 incident deaths when changing reporting practices to include \"probable" deaths as well as confirmed deaths due to COVID-19 on June 25th 2020. However, many other states have at least one outlying value, usually due to backlog reporting, where a large number of deaths occurring in previous weeks are reported on a single day. We can also see from Figure \ref{fig:data} that some states do not report on particular days (usually weekends) leading to 0 incident deaths for the day. Some states also revise the cause of death, allowing for negative incident deaths.  Some states exhibit relatively regular weekly reporting cycles, with reporting dropping off significantly on the weekends. This effect is most pronounced at the aggregate U.S. level, which shows a clear weekly cycle in reporting.

We made probabilistic forecasts  for 1-4 week ahead incident and cumulative deaths for all geographies. An individual forecast distribution is represented by a set of quantiles for incident and cumulative deaths from 1-4 weeks ahead. The quantiles used are $\mathbb{Q} = {.05,.10,....,.90,.95} \cup {.01,.99}$, with the median ($.5$ quantile) representing the point-forecast. Forecasts were made on Monday evenings and therefore use the incident data up until the Sunday before. A one-week ahead target corresponds to the following Saturday. A two-week ahead corresponds two the next Saturday and so on. Forecasts were evaluated using mean-absolute-error (MAE) and empirical coverage probability at the 95\% level. This allows us to assses both point prediction and calibration of forecasts.

 \begin{figure}
     \centering
     \includegraphics[scale=.14]{data_plot.png}
     \caption{Deaths by state for any state with over 50 incident deaths for a given day. Notice the large variability in incident reporting. There are significant data-dumps, with New Jersey reporting over 1500 backlogged deaths when switching to include "probable" deaths. In some states, there appears to be a weekly cycle where deaths are under-reported on the weekend. This is especially pronounced for the U.S. as a whole. We can also see that there are some negative incident deaths, where data are revised to account for deaths that were incorrectly attributed to COVID. }
     \label{fig:data}
 \end{figure}

@@

In a given time-step (e.g. one day), each member of the population of a single geography belongs to one of the following mutually exhaustive compartments:
 Susceptible $S$, Exposed but not yet infectious $E$, Infectious $I$, Recovered $R$, hospitalized before death $D_1$ , and  deceased $D_2$ (Figure \ref{fig:seird}). Here we assume everyone who is hospitalized will eventually become deceased in order to separate the rate into both a case fatality ratio (CFR) parameter as well as a time from symptoms to death parameter, which both have prior estimates from the literature \cite{russell2020estimating}. We omit an explicit hospitalization compartment since the available hospitalization data is highly variable by state and suffers even more reporting issues than case data.  For simplicity, we assume a closed population of size $N$. 
The following parameters govern how members of the population move between compartments:  
\begin{packed_item}
\item $\beta(t)$: transmission rate, which we allow to vary by time $t$
\item $\sigma$: rate of transition from the exposed state $E$ to infectious state $I$; i.e., $1/\sigma$ is the expected duration of the time between exposure and symptom onset. 
\item $\gamma$: rate of transition from  the infectious state $I$ to no longer being infectious (either to state $D_1$ or $R$); i.e., $1/\gamma$ is the expected duration of the infectious period
\item $\rho$: fatality rate (i.e., probability of transitioning from I to $D_1$ instead of I to R)
\item $\lambda$: rate of transition from $D_1$ to $D_2$ 
(i.e., the inverse of expected number of days in $D_1$ compartment before death) 
\end{packed_item}
For a given time-step $t$, the following differential equations describe the changes in each compartment:

@@

This model assumes that forecasts are made on the current level of interventions because $\mathbb{E}[log(\beta(t+1))] = log(\beta(t))$. That is, the expected value of a random walk in the forecasting stage is simply $\beta(t)$ at the last observed value of $t$.

@@

 The observed data used to fit the model is based on time-series data of incident confirmed cases $Cases_{t}$ and incident recorded deaths $Deaths_{t}$. 
For a given state and day, the change in the  confirmed cases and reported deaths are subset of the cumulative number of new infections $C(t)$ and cumulative number of deaths $D2(t)$, respectively. Therefore, we introduce two additional parameters for the case and death deviation model of cases $p_c$ and the case and death deviation model of deaths $p_d$. For both, we set fairly flat priors to reflect these parameters are poorly determined from observed data.

In more detail, $p_c$ is the probability that an infected person receives a positive test result and is reported by a state or local health authority as a case. We assume its prior distribution is given by  $p \sim \Beta(15, 35)$, such that  $\E[p_c] = 0.3$  with 90\% probability between \[ .22,0.38\]. This means that we expect 30\% of cases to be detected initially, as suggested by the literature \cite{midas}. However, we also allow this to vary by time.

@@

We also assume the probability that a COVID-19 death is reported $p_d$ has a prior distribution given by  $p_d \sim \Beta(90, 10)$. This prior satisfies $\E[p_d] = 0.9$ with concentration 100. That is, we assume that deaths due to COVID-19 are most often correctly reported \cite{weinberger2020estimation}.

@@

 This process ensured that the observed data does not contain any identifiable outliers (meaning documented by outside sources). In real-time this is necessary to avoid drastic over-predictions caused by data dumps. For example, New Jersey reported nearly 1,600 daily deaths as it switched from reporting only confirmed deaths from COVID-19, to confirmed and probable on 2020-06-25. This would have caused a drastic increase in predictions if not properly identified as data dump. On 2020-07-07 Texas removed 3,000 confirmed cases when they discovered the reported cases were a result of anitgen testing, which were not considered reportable. 
% \begin{itemize}
% \item Real-time forecasting evaluation for 14 weeks starting April 20th 2020. 
% \item Forecasts submitted every Monday using incident data up until Sunday
% \item Cumulative forecasts generated by aggregation of incident forecasts
% \item 1-4 week ahead targets generated from 28 day ahead predictions 
% \end{itemize}

 We compare the results against the COVID-19 Forecast Hub baseline model. This is a purely statistical model that uses historical COVID-19 deaths only.

\begin{eqnarray}
\text{Deaths}_t &\sim& N(\text{Deaths}_{t-1},\hat{\sigma}^2)\\
\hat{\sigma} &\sim& not sure
\end{eqnarray}

This model is fit state by state. This is similar to an auto-regressive order 1 model, a common time series baseline model. However, here the variance is determined from historical changes in deaths.

@@

 \begin{itemize}
 \item \textbf{MechBayes Full} Mech Bayes as of model version 3. That is, a model using negative binomial observation noise as well as a time-varying random walk, using a joint likelihood over cases and deaths.

@@

 We also fix all non-model component variation. That is, we average over the last 10 days of $\beta(t)$ when forecasting, as well as manually redistributing data dumps. This ensures that the comparison is only on model components, and not on data discrepancies. Note that we do not include a model without a time varying transmissibility parameter. This is because such a model would assume no interventions were put in place, which clearly violates the data-generating process. Previous Covid-19 modeling attempts have established that time-varying transmissibility is essential \cite{pei2020differential} \cite{abbott2020estimating}\cite{flaxman2020estimating} \cite{smirnova2019forecasting}.

@@

\begin{subfigure}{1\textwidth}
  \centering
    \includegraphics[scale=.15]{detection_plot.png}
    \caption{Time-varying case and death deviation model ($p_{c,t}$)}
\end{subfigure} 
 \caption{  A) Time varying beta parameter for all states with a day exceeding 50 incident deaths, including the U.S. by 2020-07-26. We can clearly see that the random walk is able to non-parametrically account for the non-pharmaceutical interventions retrospectively, with transmission dropping in states like New York quite rapidly. While it may seem that many states follow the same trend, this is slightly misleading since the prior pulls $\beta(t)$ upward for $t$ close to 0 in states that did not experience community transmission in March.  B) Time varying case and death deviation model for example states. The time varying detection random walk is able account for the increase in testing and other reporting anomolies. There is remarkable similarity among the estimated case and death deviation model trajectories between states.}
\label{fig:model_details}
\end{figure}

We can see from Figure \ref{fig:fit_and_forecast_results} that MechBayes version 3 is able to accurately model the observed data. The model is able to adapt to highly variable incident death reporting, variable transmission rates, and overall heterogeneity of incidence curves. The model is also able capture the uncertainty of the differential equation parameters well enough to produce well calibrated prediction intervals. Figure \ref{fig:fit_and_forecast_results}  shows prediction intervals at the 95\% level, with 92.3\% of observations falling within the bounds for each state. However, we can also see that in some states, such as California and Florida, the model is biased high, with all observations outside of the 95\% prediction interval falling below. Figure \ref{fig:fit_and_forecast_results}  also shows 4 weeks of daily forecasts, along with the daily observed incidence for 1 week out. We can see that the predictions are tracking the data even under the weekly reporting cycles.

We can see from Figure \ref{fig:model_details} that MechBayes is able to learn to adapt to the evolving pandemic situation. Panel A shows the time-varying transmission $\beta(t)$ for three example states, CA, FL, and NY as well as the U.S. We can see that $\beta(0)$ is centered on our prior but as data comes in, the estimate increases. This is especially true in NY, where the epidemic took off quickly in March. However, the model is then able to adjust to the varying levels of non-pharamceutical interventions present in each of the three states as well as across the U.S. This radically reduces the transmissibility parameter by the first week of April 2020. This is consistent with a peak in overall deaths two weeks later in mid-April 2020. There seems to be some estimation issues at the boundary of $\beta(t)$ where the number of cases does not match the number of deaths, since the cases have not yet converted to deaths. This results in the model underestimating transmissibility to reduce the flow through the compartments.

We can also see from Figure \ref{fig:model_details}  that MechBayes is able to account for the drastic increase in testing that has occurred across the U.S. since March 2020. Panel B also shows that our prior estimate of 30\% of cases being detected, may have been too high, as all regions show a dip in case and death deviation model before climbing again. Note that interpreting this strictly as time-varying detection is obscuring the fact that this parameter $p_{t,d}$ can soak-up any excess variation beyond the ability of cases and the case-fatility ratio to explain the number of deaths. That is, the time-varying case and death deviation model is able to "de-couple" cases and deaths beyond the case-fataility ratio regardless of the underlying reason (whether that be an increase in testing or shifting age distribution of cases). Thus, interpretation of this parameter as a strict mapping to testing is incorrect. As a forecasting model, we only need the ability to non-parametrically model deviations and reporting issues in cases.

@@

\caption{Scores from COVID-19 Forecast Hub broken down by target and forecast week. We use the relative MAE metric which takes the MAE of MechBayes model and divides by the MAE of the baseline model. Here we can see that the MechBayes model improves in both MAE and coverage probability over time, consistently beating the baseline model in the month of July 2020. We can see that as the horizon increases from 1 to 4 weeks ahead, MAE, reflecting an increase in difficulty of forecasting further ahead in time. Note that targets averaging over forecast week include an average over all model versions. We also see a uniform improvement in coverage probability when broken down by timezero and target. Here the horizontal line represents 95\% coverage interval. }
\label{fig:covidhub_appendix}
\end{figure}

@@

We begin by breaking down the results by week the forecast was made (forecast week) and averaging over both region and target. As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes outperformed the COVID-19 Forecast Hub baseline on MAE and coverage probability when broken down by forecast week.  As we can see from \ref{fig:data}, in weeks with the largest increase in deaths (mostly during the month of May) MechBayes significantly outperformed the baseline model. However, in weeks with a small increase or a decrease in deaths, the scores were much closer. This suggests that MechBayes performs well where it counts, when the epidemic is taking off nearly exponentially.

We also break down the results by geographical region, as seen in Figure \ref{fig:covidhub} (C,D). Note that this break-down is averaged over target (1-4 week ahead), but also forecast week, meaning that we average over each of the model versions. We feel this is important, as it reflects the real-time accuracy of our evolving model efforts, instead of cherry-picking the best model. However, we still see consistent improvements in MAE under MechBayes when broken down by region. We see the largest improvements in the regions where the baseline model had the highest MAE, meaning that MechBayes improves forecasts in regions with more significant viral activity.

We break down the results by target by averaging over region and forecast week  \ref{fig:covidhub} (E,F) Here we can see uniform improvement over the baseline model by MechBayes in terms of MAE and WIS. We can also see that the MAE increase as horizon increase, which is to be expected. We can also see that incident MAE is lower than cumulative MAE, which is again to be expected due to the lower absolute numbers of incident deaths.

Finally, we include a formal test of the difference in MAE and WIS to demonstrate statistically significant advantages in using MechBayes over the baseline model. In order to do this, we just a random effects regression model of the form,

@@

where $mb$ is an indicator for the MechBayes model, $t$ is timezero, $r$ is region, and $h$ is target horizon (1-4 week ahead). We chose this model because it explains the variation in MAE by model and horizon while allowing varying baseline MAE values by region. Here, variation over time in MAE within a specific region is explained by differences in model performance This leads to the following coefficient estimates for the fixed effects,

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & Estimate & Std. Error & df & t value & Pr($>$$|$t$|$) \\ 
  \hline
$\beta_0$  & 2.47 & 0.17 & 54.87 & 14.97 & 0.00 \\ 
   $\beta_{h_1}$ & 0.30 & 0.06 & 4643.00 & 5.18 & 0.00 \\ 
 $\beta_{h_2} $ & 0.44 & 0.06 & 4643.00 & 7.56 & 0.00 \\ 
  $\beta_{h_3}$ & 0.59 & 0.06 & 4643.00 & 9.79 & 0.00 \\ 
  $\beta_{h_1 mb}$  & -0.19 & 0.06 & 4643.00 & -3.19 & 0.00 \\ 
  $\beta_{h_2 mb}$ & -0.29 & 0.06 & 4643.00 & -4.87 & 0.00 \\ 
   $\beta_{h_3mb}$ & -0.23 & 0.06 & 4643.00 & -3.94 & 0.00 \\ 
 $\beta_{h_4mb}$ & -0.29 & 0.06 & 4643.00 & -4.72 & 0.00 \\ 
   \hline
\end{tabular}
\caption{Coefficient estimates and t-values for MAE evaluation model. We can see that MechBayes performs statistically significantly better than the baseline model for 2-4 weeks ahead. The performance increase does not seem to follow a particular pattern.}
\end{table}

>>>>

The COVID-19 pandemic emerged in late December 2019. In the first six months of the global outbreak, the US reported more cases and deaths than any other country in the world. Effective modeling of the course of the pandemic can help assist with public health resource planning, intervention efforts, and vaccine clinical trials. However, data quality during a pandemic suffers from under-reporting, delayed reporting due to reporting infrastructure issues, and limited testing. Traditionally, compartmental models have been used to model emerging pandemics. However, the basic compartmental framework is difficult to operationalize in a complex data environment. We propose a novel Bayesian compartmental model (MechBayes) that builds upon the classic compartmental framework of susceptible-exposed-infected-recovered (SEIR) model to operationalize COVID-19 forecasting in real time. This includes non-parametric modeling of varying transmission rates, non-parametric modeling of case and death discrepancies due to testing and reporting issues, and finally joint observation likelihood on new case counts and new deaths. The model has been used to submit forecasts to the US Centers for Disease Control, through the COVID19-ForecsatHub repository organized by the Reich Lab. We examine the performance relative to a baseline model in addition to performing an ablation test of our extensions to the classic SEIR models. We demonstrate a significant gain in both point and probabilistic forecast scoring measures using MechBayes.
%\begin{itemize}
%\item Covid has affected XX people
%\item Forecasts useful for public health resource planning, intervention planning (vaccine trials), and disease burden
%\item Basic Mechanistic Models unable to capture complexities of real world disease epidemics due to 
%\begin{itemize}
%\item Complexities of interventions 
%\item Issues with testing and reporting
%\end{itemize}
%\item We propose a novel forecasting algorithm to overcome 
%\begin{itemize}
%\item Under-reporting of cases
%\item Time-varying interventions
%\end{itemize}
%\item Bayesian end to end estimation using both cases and deaths in numpyro
%\end{itemize}

@@

In this work, we introduce a novel operational forecast model based on mechanistic foundations and tailored to the particular needs and data availability of COVID-19. These include, but are not limited to, severe under reporting of cases due to low testing rates especially in mild or asymptomatic cases, time-varying testing rates, delayed reporting, and both the addition and removal of control measures such as social distancing, lockdown, and mask use. We accomplish this by modeling transmission and deviations from cases and deaths (beyond the case fatality ratio) non-parametrically. Our observation model is able to handle case data which is reported as new cases, instead of prevalent, and new deaths, instead of cumulative. We choose a Bayesian framework that allows for uncertainty in the epidemiological parameters that are unidentifiable from the data, introducing flexibility suited to forecasting. Implementation in a Bayesian framework also allows for use of a cutting edge probabilistic programming framework for speed and rapid model development.

Our main goal in developing MechBayes is to forecast observed incident deaths, and because we have limited historical data on COVID-19, we think compartmental models are among the most parsimonious models available for forecasting. Our focus is not on identifying internal parameters of the model, many of which are poorly determined or not identifiable from the available data. We are willing to accept biologically unrealistic parameter values or priors for latent variables; but not ones that lead to gross pathologies during inference and forecasting. The non-parametric extensions to the underlying SEIR model also mean that any inference on epidemiological parameters should be carried out very carefully. We distinguish this from scenario projection models, which require well identified epidemiological parameters that can be set to counterfactual values under different scenarios, such as an increase or decrease in intervention levels.

 We demonstrate the success of the model in both forecast submissions to the US Centers for Disease Control via the COVID-19 Forecast Hub as well as an ablation model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model. In what follows we first describe the available data and forecast submission infrastructure, outline the basic SEIR compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.

@@

\section{Data}
	In this analysis we used confirmed case counts and deaths as reported by the Johns Hopkins University Center for Systems Science and Engineering \cite{dong2020interactive}. This a time series dataset which we truncate to begin March 1st 2020 to August 1st 2020 and captures all 50 states, as well as Guam, Puerto Rico, American Samoa, District of Columbia, Northern Mariana Islands and U.S. Virgin Islands. As noted in \cite{krantz2020level}, COVID-19 cases are under-reported, with the fraction of all infections reported as cases for the U.S. estimated at 20-30\% \cite{russel2020using}. 
		There are large discrepancies in reporting practices across the states (Figure \ref{fig:data}). For example, New Jersey reported an additional 1600 incident deaths when changing reporting practices to include ``probable" deaths as well as confirmed deaths due to COVID-19 on June 25th 2020. However, many other states have at least one outlying value, usually due to delayed reporting, where a large number of deaths occurring in previous weeks are reported on a single day. We can also see from Figure \ref{fig:data} that some states do not report on particular days (usually weekends) leading to 0 incident deaths for the day. Some states also revise the cause of death, allowing for negative incident deaths.  Some states exhibit relatively regular weekly reporting cycles, with reporting dropping off significantly on the weekends. This effect is most pronounced at the aggregate U.S. level, which shows a clear weekly cycle in reporting.

We made probabilistic forecasts  for 1-4 week ahead incident and cumulative deaths for all geographies. An individual forecast distribution is represented by a set of quantiles for incident and cumulative deaths from 1-4 weeks ahead. Since our model produced cumulative forecasts by aggregating incident, we choose to only evaluate incident. The quantiles used are $\mathbb{Q} = {.05,.10,....,.90,.95} \cup {.01,.99}$, with the median ($.5$ quantile) representing the point-forecast. Forecasts were made on Monday evenings and therefore use the incident data up until the Sunday before. A one-week ahead target corresponds to the following Saturday. A two-week ahead corresponds two the next Saturday and so on. Forecasts were evaluated using mean-absolute-error (MAE) and empirical coverage probability at the 95\% level. This allows us to assses both point prediction and calibration of forecasts.

 \begin{figure}
     \centering
     \includegraphics[scale=.14]{data_plot.png}
     \caption{Deaths by state for any state with over 50 incident deaths for a given day. Notice the large variability in incident reporting. There are significant delayed reporting, with New Jersey reporting over 1500 backlogged deaths when switching to include ``probable" deaths. In some states, there appears to be a weekly cycle where deaths are under-reported on the weekend. This is especially pronounced for the U.S. as a whole. We can also see that there are some negative incident deaths, where data are revised to account for deaths that were incorrectly attributed to COVID-19. }
     \label{fig:data}
 \end{figure}

@@

In a given time-step (e.g. one day), each member of the population of a single geography belongs to one of the following mutually exhaustive compartments:
 susceptible $S$, exposed but not yet infectious $E$, infectious $I$, recovered $R$, hospitalized before death $D_1$, and deceased $D_2$ (Figure \ref{fig:seird}). Here we assume everyone who is hospitalized will eventually become deceased in order to separate the rate of flow into both a case fatality ratio (CFR) parameter as well as a time from symptoms to death parameter, which both have prior estimates from the literature \cite{russell2020estimating}. We omit an explicit hospitalization compartment since the available hospitalization data is highly variable by state and suffers even more reporting issues than case data.  For simplicity, we assume a closed population of size $N$. 
The following parameters govern how members of the population move between compartments:  
\begin{packed_item}
\item $\beta(t)$: transmission rate, which we allow to vary by time $t$
\item $\sigma$: rate of transition from the exposed state $E$ to infectious state $I$; i.e., $1/\sigma$ is the expected duration of the time between exposure and symptom onset. 
\item $\gamma$: rate of transition from  the infectious state $I$ to no longer being infectious (either to state $D_1$ or $R$); i.e., $1/\gamma$ is the expected duration of the infectious period
\item $\rho$: fatality rate (i.e., probability of transitioning from I to $D_1$ instead of I to R)
\item $\lambda$: rate of transition from $D_1$ to $D_2$ 
(i.e., the inverse of expected number of days in $D_1$ compartment before death) 
\end{packed_item}
For a given time-step $t$, the following differential equations describe the changes in each compartment:

@@

This model assumes that forecasts are made on the current level of interventions because $\mathbb{E}[log(\beta(t+1))] = log(\beta(t))$. That is, the expected value of a random walk in the forecasting stage is the estimated value at the final time point of available data.

@@

 The observed data used to fit the model is based on time-series data of incident confirmed cases $Cases_{t}$ and incident recorded deaths $Deaths_{t}$. 
For a given state and day, the change in the  confirmed cases and reported deaths are subset of the cumulative number of new infections $C(t)$ and cumulative number of deaths $D2(t)$, respectively. To handle this, we introduce two additional parameters. First, we introduce a case and death deviation, beyond the case fatality ration, of cases $p_c$ and the case and death deviation model of deaths $p_d$. For both, we set fairly flat priors to reflect these parameters are poorly determined from observed data.

In more detail, $p_c$ can be thought of as an aggregate probability of a case being detected and then flowing through the compartments. We assume its prior distribution is given by  $p \sim \Beta(15, 35)$, such that  $\E[p_c] = 0.3$  with 90\% probability between $0.22,0.38$. This means that we expect 30\% of cases to be detected initially, as suggested by the literature \cite{midas}. However, we also allow this to vary by time. We do not intend for this to be interpretable as purely reflecting testing, but rather an aggregate measure of testing, reporting issues, and general departure from our prior estimate of the case fatality ratio.

@@

We also assume the probability that a COVID-19 death is reported $p_d$ has a prior distribution given by  $p_d \sim \Beta(90, 10)$. This prior satisfies $\E[p_d] = 0.9$ with 90\% probability between 0.89,0.92. That is, we assume that deaths due to COVID-19 are most often correctly reported \cite{weinberger2020estimation}.

@@

 This process ensured that the observed data does not contain any identifiable outliers (meaning documented by outside sources). In real-time this is necessary to avoid drastic over-predictions caused by delayed reporting. For example, New Jersey reported nearly 1,600 daily deaths as it switched from reporting only confirmed deaths from COVID-19, to confirmed and probable on 2020-06-25. This would have caused a drastic increase in predictions if not properly identified as data dump. On 2020-07-07 Texas removed 3,000 confirmed cases when they discovered the reported cases were a result of anitgen testing, which were not considered reportable. 
% \begin{itemize}
% \item Real-time forecasting evaluation for 14 weeks starting April 20th 2020. 
% \item Forecasts submitted every Monday using incident data up until Sunday
% \item Cumulative forecasts generated by aggregation of incident forecasts
% \item 1-4 week ahead targets generated from 28 day ahead predictions 
% \end{itemize}

 We compare the results against the COVID-19 Forecast Hub baseline model. This is a purely statistical model that uses historical COVID-19 deaths only. At all forecast horizons, the median of the forecast distribution from the baseline model is equal to the most recent reported incidence. The model obtains a non-parametric distribution around this median by adding and subtracting past observed differences in incidence from one week to the next for a specific location. This model is fit state by state.

@@

 \begin{itemize}
 \item \textbf{MechBayes Full} Mech Bayes as submitted to the CDC. That is, a model using negative binomial observation noise as well as a time-varying random walk, using a joint likelihood over cases and deaths.

@@

 We also fix all non-model component variation. That is, we average over the last 10 days of $\beta(t)$ when forecasting, as well as manually redistributing delayed reporting. This ensures that the comparison is only on model components, and not on data discrepancies. Note that we do not include a model without a time varying transmissibility parameter. This is because such a model would assume no interventions were put in place, which clearly violates the data-generating process. Previous Covid-19 modeling attempts have established that time-varying transmissibility is essential \cite{pei2020differential} \cite{abbott2020estimating}\cite{flaxman2020estimating} \cite{smirnova2019forecasting}.

@@

\begin{subfigure}{1\textwidth}
  \centering
    \includegraphics[scale=.15]{detection_plot.png}
    \caption{Time-varying case and death deviation model ($p_{c,t}$)}
\end{subfigure} 
 \caption{  A) Time varying beta parameter for all states with a day exceeding 50 incident deaths, including the U.S. by 2020-07-26. We can clearly see that the random walk is able to account for a composite of intervention efforts. B) Time varying case and death deviation model for example states. The time varying case and death deviation model is able account for the increase in testing and other reporting anomolies.}
\label{fig:model_details}
\end{figure}

\section{Real-time Forecasting Results}

We can see from Figure \ref{fig:fit_and_forecast_results} that MechBayes is able to accurately model the observed data. The model is able to adapt to highly variable incident death reporting, variable transmission rates, and overall heterogeneity of incidence curves. The model is also able capture the uncertainty of the differential equation parameters well enough to produce well calibrated prediction intervals. Figure \ref{fig:fit_and_forecast_results}  shows prediction intervals at the 95\% level, with 92.3\% of observations falling within the bounds for each state. However, we can also see that in some states, such as California and Florida, the model is biased high, with all observations outside of the 95\% prediction interval falling below. Figure \ref{fig:fit_and_forecast_results}  also shows 4 weeks of daily forecasts, along with the daily observed incidence for 1 week out. We can see that the predictions are tracking the data even under the weekly reporting cycles.

We can see from Figure \ref{fig:model_details} that MechBayes is able to learn to adapt to the evolving pandemic situation. Panel A shows the time-varying transmission $\beta(t)$ for 20 states with the highest total deaths. We can see that $\beta(0)$ is centered on our prior but as data comes in, the estimate increases. This is especially true in NY, where the epidemic took off quickly in March. However, the model is then able to adjust to the varying levels of interventions present in each state. This radically reduces the transmissibility parameter by the first week of April 2020 for almost all states. This is consistent with a peak in overall deaths two weeks later in mid-April 2020. There seems to be some estimation issues at the boundary of $\beta(t)$ where the number of cases does not match the number of deaths, since the cases have not yet converted to deaths. This results in the model underestimating transmissibility to reduce the flow through the compartments.

We can also see from Figure \ref{fig:model_details}  that MechBayes is able to account for the drastic increase in testing that has occurred across the U.S. since March 2020. Panel B also shows that our prior estimate of 30\% of cases being detected, may have been too high, as all regions show a dip in case and death deviation model before climbing again.

%Note that interpreting this strictly as time-varying detection is obscuring the fact that this parameter $p_{t,d}$ can soak-up any excess variation beyond the ability of cases and the case-fatility ratio to explain the number of deaths. That is, the time-varying case and death deviation model is able to "de-couple" cases and deaths beyond the case-fataility ratio regardless of the underlying reason (whether that be an increase in testing or shifting age distribution of cases). Thus, interpretation of this parameter as a strict mapping to testing is incorrect. As a forecasting model, we only need the ability to non-parametrically model deviations and reporting issues in cases.

@@

\caption{Scores from COVID-19 Forecast Hub broken down by target and forecast week. MechBayes improves in average MAE in 9 out of 12 timepoints. MechBayes improves in MAE across all targets. We can see that as the horizon increases from 1 to 4 weeks ahead, MAE, reflecting an increase in difficulty of forecasting further ahead in time. We also see a uniform improvement in coverage probability when broken down by timezero and target. Here the horizontal line represents 95\% coverage interval. }
\label{fig:covidhub}
\end{figure}

@@

We begin by breaking down the results by week the forecast was made (forecast week) and averaging over both region and target. As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes outperformed the COVID-19 Forecast Hub baseline on MAE and coverage probability when broken down by forecast week. However, the improvements broken down by time are small, and improves in 9 out of 12 evaluation weeks.  As we can see from \ref{fig:data}, in weeks with the largest increase in deaths (mostly during the month of May) MechBayes significantly outperformed the baseline model. However, in weeks with a small increase or a decrease in deaths, the scores were much closer. This suggests that MechBayes performs well where it counts, when the epidemic is taking off nearly exponentially.

We also break down the results by geographical region, as seen in Figure \ref{fig:covidhub} (C,D). Here we see consistent improvements in MAE under MechBayes for regions with high total death counts with an improvement in the average MAE in 16 out of the 20 states with the highest total death count. Out of the states with the 10 highest death counts, only in California did the baseline model outperform MechBayes on the average MAE. However, the results are more mixed for states with lower death counts.

We break down the results by target by averaging over region and forecast week  \ref{fig:covidhub} (E,F) Here we can see uniform improvement over the baseline model by MechBayes in terms of MAE and coverage probability. We can also see that the MAE increase as horizon increase, which is to be expected. We can also see that incident MAE is lower than cumulative MAE, which is again to be expected due to the lower absolute numbers of incident deaths.

Finally, we include a formal test of the difference in MAE to demonstrate statistically significant advantages in using MechBayes over the baseline model. In order to do this, we just a random effects regression model of the form,

@@

where $mb$ is an indicator for the MechBayes model, $t$ is timezero, $r$ is region, and $h$ is target horizon (1-4 week ahead). We chose this model because it explains the variation in MAE by model and horizon while allowing varying baseline MAE values by region. Here, variation over time in MAE within a specific region is explained by differences in model performance This leads to the following coefficient estimates for the fixed effects.

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & Estimate & Std. Error & df & t value & Pr($>$$|$t$|$) \\ 
  \hline
$\beta_0$  & 2.47 & 0.17 & 54.87 & 14.97 & 0.00 \\ 
   $\beta_{h_1}$ & 0.30 & 0.06 & 4643.00 & 5.18 & 0.00 \\ 
 $\beta_{h_2} $ & 0.44 & 0.06 & 4643.00 & 7.56 & 0.00 \\ 
  $\beta_{h_3}$ & 0.59 & 0.06 & 4643.00 & 9.79 & 0.00 \\ 
  $\beta_{h_1 mb}$  & -0.19 & 0.06 & 4643.00 & -3.19 & 0.00 \\ 
  $\beta_{h_2 mb}$ & -0.29 & 0.06 & 4643.00 & -4.87 & 0.00 \\ 
   $\beta_{h_3mb}$ & -0.23 & 0.06 & 4643.00 & -3.94 & 0.00 \\ 
 $\beta_{h_4mb}$ & -0.29 & 0.06 & 4643.00 & -4.72 & 0.00 \\ 
   \hline
\end{tabular}
\caption{Coefficient estimates and t-values for MAE evaluation model. We can see that MechBayes performs statistically significantly better than the baseline model for 1-4 weeks ahead. The performance increase seems to grow as horizon increases.}
\end{table}
